---
title: "Marine Spatial PLanning"
author: "Moegamad Uzair Jack"
date: "2025-06-02"
output:
  html_document: default
  word_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading packages

```{r}
library(tidyverse)
library(ggpubr)
library(vegan)
library(readxl)
```

## Reading File
Reading csv as a data frame

```{r}
EF <- read.csv('EstuaryFishSurveyData.csv')
class(EF)
```
Replace NA values with 0

```{r}
EF[is.na(EF)] <- 0
```


## Alpha Diversity
Creating a separate column

```{r}
EF$alpha <- specnumber(EF[4: 148])
```

Plotting alpha diversity (species number) per estuary

```{r}
EF$Estuary <- factor(EF$Estuary, levels = unique(EF$Estuary)) # Ensure Estuary is a factor for ordered x-axis labels

g1 <- ggplot(EF, aes(x = Estuary, y = alpha)) +
  geom_bar(stat = "identity", fill = "grey") +
  theme_minimal() +
  labs(
    title = "Alpha Diversity Across Estuary",
    x = "Estuary",
    y = "Alpha diversity"
  ) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 5),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5),
    panel.grid.major.x = element_blank(),
    panel.grid.minor = element_blank()
        )
g1
```

## SW Diversity

```{r}
EF$SW <- diversity(EF[4:148], index = "shannon")
```

Creating kmEast column
```{r}
EF$kmEast <- 3000 - EF$kmWest
```
## Plots
Plotting alpha diveristy vs. km East

```{r}
g2 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Alpha Diversity vs. km East",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
g2
```


Plotting Shannon–Wiener Diversity vs. km East

```{r}
g3 <- ggplot(EF, aes(x = kmEast, y = SW)) +
  geom_point(color = "steelblue", size = 2) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Shannon–Wiener Diversity vs. km East",
    x = "km East",
    y = "Shannon–Wiener Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10)
  )
g3
```

## Data frame manipulation
Reorder data frame

```{r}
EF <- EF[order(-EF$alpha), ]

```

Creating new variable

```{r}
EF$AlphaList <- 0
EF$AlphaList[1:20] <- 1
```

## Plots
Plot top 20 in RED and the rest in BLACK
 Plot the alpha diversity against km East with colour factor based on the alpha rank. 
 Here the  top 20 are represented in red and the rest in black.

```{r}

g4 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(AlphaList)), size = 2) +
    scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top 20 estuaries"),
    name = "Alpha Rank"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Alpha Diversity vs. km East",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g4

```

## Alpha Method
Now we need to list the top 20 esturaries and save it into a variable

```{r}
AlphaList <- EF[1:20, 1]
AlphaList
```

We want the list and the number of species.
0 represents the species not in the list; 1 represents the species that is in the list.
```{r}
AlphaSpp <- specnumber(EF[,4:148] , EF$AlphaList)
AlphaSpp
```
There are 116 species in the list and 145 species in total. So, we calculate the  percentage of species within the list.
```{r}
AlphaPerc <- (116/145) *100
AlphaPerc
```
## Change the metric
Now do the same for the SW index. Start by reordering the dataframe

```{r}
EF <- EF[order(-EF$SW), ]

```
Creating new variable

```{r}
EF$SW_list <- 0
EF$SW_list[1:20] <- 1
```

Plot top 20 in RED and the rest in BLACK
Plot the SW against km East with colour factor based on the SW rank. 
Here the  top 20 are represented in red and the rest in black.
Keep alpha to compare between the 2.

```{r}

g5 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(SW_list)), size = 2) +
    scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top 20 estuaries"),
    name = "SW Rank"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "SW index vs. km East",
    x = "km East",
    y = "SW"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g5

```
## SW Method
Now we need to list the top 20 esturaries and save it into a variable

```{r}
SW_list <- EF[1:20, 1]
SW_list
```

We want the list and the number of species.
0 represents the species not in the list; 1 represents the species that is in the list.
```{r}
SW_spp <- specnumber(EF[,4:148] , EF$SW_list)
SW_spp
```
There are 110 species in the list and 145 species in total. So, we calculate the  percentage of species within the list.
```{r}
SW_perc <- (110/145) *100
SW_perc
```

## Biogeographic Zonation
Use specnumber to calculate the species number per biogeographic zone (BZ)
```{r}
specnumber(EF[,4:148], EF$BZ)
```
In each of the 3 zones, choose the most diverse estuary in each  zone.
Choose 6 estuaries from each zone (2 spare).
6 from W; 7 from S; 7 from E.
Start ordering the data frame by BZ then by alpha.

```{r}
EF <- EF[order(EF$BZ, -EF$alpha), ]
head(EF)
```
Creating a list

```{r}

EF$BZ_list <- 0

top_n <- c(W = 6, S = 7, E = 7)

for (z in names(top_n)) {
  zone_rows <- which(EF$BZ == z)
  EF$BZ_list[head(zone_rows, top_n[z])] <- 1
}

```
Creating the plot
```{r}
g6 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(BZ_list)), size = 2) +
    scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top estuaries"),
    name = "Alpha Rank"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "BZ List",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g6
```
We want the list and the number of species.
0 represents the species not in the list; 1 represents the species that is in the list.
```{r}
BZ_spp <- specnumber(EF[, 4:148], EF$BZ_list)
BZ_spp
```
There are 122 species in the list and 145 species in total. So, we calculate the  percentage of species within the list.
```{r}
BZ_perc <- (122/145) *100
BZ_perc
```
## Dendrogam
We have things called biogeographical zones, each with their own sets of species. When 2 zones come together (overlap), that is where you get the most species. But that is not good to protect. The edges have more diversity but we need to protect the centre. The edges have high diversity and are not good for protection because they are a mix and not  atrue measure of diversity.
We do multivariate. We find if the ecosystem falls into groups and then we identify the estuaries that fall into these ecossytem groups. To do this, we make use of a dendrogram.

Create a dendrogram based on species composition. 'vegan' plays a key role in this process.
Pairwise comparisons between each estuary to develop an index of similarity in species composition.

```{r}
EF <- subset(EF, alpha > 0)  #subset the data due to the 0s

EF_deco <- decostand(EF[,4:148], method = 'total')
```


## Bray-Curtis
Bray-Curtis most commonly used index for similarity. Especially when there are many 0s, Bray-CUrtis is a good fit.

```{r}
EF_dist <- vegdist(EF_deco, method =  'bray')
```
## H-clust and V-clust
H-clust is horizontal and v-clust is vertical.

```{r}

EF_clust <- hclust(EF_dist, method = "average") # run clustering

plot(
  EF_clust,
  labels = EF$Estuary,
  hang = -1,
  cex = 0.6,
  main = NULL,
  ylab = "Dissimilarity",
  sub = "",
  col = "black"
)
```
Reduce the tree
```{r}
slice <- cutree(EF_clust,
                h= 0.8)
slice

EF$Slice80 <- slice #adding into data frame
```
Ordering data by slice then alpha
```{r}
EF <- EF[order(EF$Slice80, -EF$alpha), ]
```
#
Creating list and score the top 2 in each Slice80 group as 1 and the rest as 0
```{r}
# Initialize the Slice80_list column to 0
EF$Slice80_list <- 0

# Loop through each unique Slice80 group
for (g in unique(EF$Slice80)) {
  
  # Find all rows in the group
  group <- which(EF$Slice80 == g)
  
  # Determine how many rows to score
  top_n <- if (g == "2") {
    4
  } else if (g == "3") {
    3
  } else {
    2
  }
  
  # Always score at least 1 row (if only 1 available)
  EF$Slice80_list[group[1:min(top_n, length(group))]] <- 1
}

```

## Plots

```{r}
g7 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(Slice80_list)), size = 2) +
    scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top estuaries"),
    name = "Alpha Rank"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "SLice 80 List",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g7
```
List the species in the groups
```{r}
Slice80_spp <- specnumber(EF[,4:148], EF$Slice80_list)
Slice80_spp
```
There are 100 species in the list and 145 species in total. So, we calculate the  percentage of species within the list.
```{r}
Slice80_perc <- (100/145) *100
Slice80_perc
```
## The Range Centre Method

Calculate the range
```{r}
# species abundance data
spp <- EF[, 4:148]

# distance from eastern boundary
coords <- EF$kmEast
```
Now we need to calculate the range for each species. We can create a function to do this.
```{r}
range_stats <- function(abundance, coords) {
  present_coords <- coords[abundance > 0]  # get kmEast values where species is present
  
  if (length(present_coords) == 0) {
    # species not present anywhere
    return(c(west_limit = NA, east_limit = NA, center = NA))
  } else {
    east <- min(present_coords)                    # furthest west point
    west <- max(present_coords)                    # furthest east point
    centre <- mean(c(east, west))                  # center of range
    
    return(c(east_limit = east, west_limit = west, centre = centre))
  }
}

# Apply the function across all species (columns)
spp_range <- t(apply(spp, 2, range_stats, coords = coords))

```
Now we just need to plop our results into a dataframe which contains the species names as well as their west_limit, east_limit and centre values.
```{r}
# convert result to a clean data frame
spp_range_list <- data.frame(
  Species = colnames(spp),  # species names from column headers
  spp_range                 # range stats: west, east, center
)

# preview first few rows of the result
head(spp_range_list)
```

Now we can plot a histogram with the frequency of range centres per 100 km.
```{r}
g8<- ggplot(spp_range_list, aes(x = centre)) +
  geom_histogram(binwidth = 100, fill = "grey", color = "black") +
  xlab("Range Centre (km East)") +
  ylab("Number of Species") +
  ggtitle("Frequency of Species Range Centres (per 100 km)") +
  theme_minimal()

g8 
```
Extract frequencies from histogram and add that to the original data frame
```{r}

EF$block <- ceiling(EF$kmEast/100)

EF$RC_list <- 1

EF <- EF[order(EF$kmEast),]

interval_table <- table(cut(spp_range_list$centre,seq(0,3000,100)))
print(order(interval_table, decreasing = TRUE))
```

```{r}
# Define blocks to score as 0
zero_blocks <- c(13, 14, 3, 9, 11, 2, 4, 5, 6, 8)

# Loop through each row
for (i in 1:nrow(EF)) {
  if (EF$block[i] %in% zero_blocks) {
    EF$RC_list[i] <- 0
  }
}

```

Subset data and remove the zeros
```{r}
EF <- subset(EF, RC_list > 0) 
```

Order data by block and alpha
```{r}
EF <- EF[order(EF$block, -EF$alpha), ]
```

Assigning max
```{r}
max <- EF$alpha[1]  # Initialize max with the first alpha value
EF$RC_list <- 0 
EF$RC_list[1] <- 1

for (j in 2:nrow(EF)) {
  if (EF$block[j] > EF$block[j-1]) {
    max <- 0
    if (EF$alpha[j] > max) {
     EF$RC_list[j] <- 1 
    max <- EF$alpha[j]
    }
  } else if (EF$block[j] == EF$block[j-1]) {
    if (EF$alpha[j] > max) {
      EF$RC_list[j] <- 1
      EF$RC_list[j-1] <- 0
      max <- EF$alpha[j]
    }
  }
}

sum(EF$RC_list)
```
List the species in the groups
```{r}
RC_spp <- specnumber(EF[,4:148], EF$RC_list)
RC_spp
```
There are 121 species in the list and 145 species in total. So, we calculate the  percentage of species within the list.
```{r}
RC_perc <- (121/145) *100
RC_perc
```
## Plots

```{r}
g9 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(RC_list)), size = 2) +
     scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top estuaries"),
    name = "Top 20 of method"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "RC List",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g9
```
## Random Method

```{r}

max <- 0
EF$rand_list <- 0
EF$rand_best_list <- 0

while (max <= 100) {
  EF$rand_list <- 0
  EF <- EF[sample(1:nrow(EF)), ]
  EF$rand_list[1:20] <- 1
   
  rand_spp <- specnumber(EF[, 4:148], EF$rand_list)
   
  if (rand_spp[2] > max) {
    EF$rand_best_list  <- EF$rand_list
      max <- rand_spp[2]
      print(max)
   }
}

cat("Final max:", max, "\n")
```

So, we calculate the  percentage of species within the list.
```{r}
rand_perc <- (113/145) *100
rand_perc
```
Order and extract the estuary names
```{r}
EF <- EF[order(-EF$rand_best_list, -EF$alpha), ]

# Extract estuary names where rand_best_list == 1
top_estuaries <- EF$Estuary[EF$rand_best_list == 1]

# View or print
top_estuaries

```
Plot
```{r}
g10 <- ggplot(EF, aes(x = kmEast, y = alpha)) +
  geom_point(aes(color = factor(rand_best_list)), size = 2) +
     scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top estuaries"),
    name = "Top 20 of random method"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Random Best List",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g10
```


## Complementary Method
Sort data frame by alpha.
Order list + Take most diverse estuary and put it in list.
Remove  spp  that occurred in that most diverse estuary.
Calculate alpha diversity again (for all estuaries).
Then  take the second on the list. 
Calculate alpha again + order list.

```{r}
# Read in the data
EF2 <- read.csv('EstuaryFishSurveyData.csv')
EF2[is.na(EF2)] <- 0
EF2$kmEast <- 3000 - EF2$kmWest

EF2$alpha <- specnumber(EF2[,4:148])

EF2$alpha_sort <- specnumber(EF2[,4:148])

EF2$comp_list <- 0
spp_count <- 0

# Loop to select 20 estuaries that contribute most unique species
for (i in 1:20) {
  EF2$alpha_sort <- specnumber(EF2[, 4:148])  # richness of remaining species per estuary
  EF2 <- EF2[order(-EF2$alpha_sort), ]            # sort by descending richness
  
  spp_count <-spp_count + EF2$alpha_sort[1]
  
  EF2$comp_list[1] <- 1                    # mark the top estuary
  
    # Remove species already found in the top estuary from all rows
  for (j in 4:148) {
      if (EF2[1,j] > 0) {
       EF2[,j] <- 0
    }
  }
}

comest <- subset(EF2, comp_list > 0)
print(comest$Estuary)


cat("Number of species:", spp_count, "\n")
comp_perc <- (spp_count / 145) * 100
cat("Percentage of species:", comp_perc, "%\n")
```
Plot
```{r}

g11<- ggplot(EF2, aes(x = kmEast, y = alpha, color = factor(comp_list))) +
  geom_point(alpha = 0.7, size = 2) +
   scale_color_manual(
    values = c("0" = "black", "1" = "red"),
    labels = c("Other estuaries", "Top estuaries"),
    name = "Top 20 of complementary method"
  ) +
  theme_minimal(base_size = 12) +
  labs(
    title = "Complementary estuaries by Alpha diversity",
    x = "km East",
    y = "Alpha Diversity"
  ) +
  theme(
    plot.title = element_text(face = "bold", hjust = 0.5),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    legend.position = "top"
  )

g11
```

```{r}
#Different approaches, different lists
EF <- EF[order(-EF$alpha),]
EF2 <- EF2[order(-EF$alpha),]
EF$con_list <- EF$AlphaList + EF$SW_list + EF$BZ_list + EF$Slice80_list + EF$RC_list + EF2$comp_list

EF <- EF[order(-EF$con),]
con_list <- EF[1:20,1]
EF$con_list <- 0
EF$con_list[1:20] <- 1

g12 <- ggplot(EF, aes(x = kmEast, y = alpha)) + 
  geom_point(size = 2,  aes(colour = as.factor(con_list))) +
  scale_color_manual( values = c("black", "red"),
                      label = c("Other estuaries", "Top 20")) +
  theme_bw() +
  theme(axis.ticks.length = unit(-0.25, "cm") ) +
  theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank()) +
  labs(  x = "km East",
         y = "Alpha Diversity",
         col = "Consensus Method Selection")
g12

```
## Consensus Method
```{r}
#slice80 spp
conN <- specnumber(EF[4:148], EF$con_list) #112 species protected

#slice80 %
con_perc <- (conN[2]/145)*100 

# Create estuary_table with Estuary column included
estuary_table <- data.frame(
  Estuary = as.character(EF$Estuary),
  Alpha = EF$AlphaList,
  SW = EF$SW_list,
  BZ = EF$BZ_list,
  Slice80 = EF$Slice80_list,
  RC = EF$RC_list,
  Comp = EF2$comp_list,
  Consensus = EF$con_list,
  stringsAsFactors = FALSE
)

# Create summary rows
species_row <- data.frame(
  Estuary = "Species Protected",
  Alpha = AlphaSpp[2],
  SW = SW_spp[2],
  BZ = BZ_spp[2],
  Slice80 = Slice80_spp[2],
  RC = RC_spp[2],
  Comp = spp_count,
  Consensus = conN[2],
  stringsAsFactors = FALSE
)

percent_row <- data.frame(
  Estuary = "Percent Protected",
  Alpha = ceiling(AlphaPerc),
  SW = ceiling(SW_perc),
  BZ = ceiling(BZ_perc),
  Slice80 = ceiling(Slice80_perc),
  RC = ceiling(RC_perc),
  Comp = ceiling(comp_perc),
  Consensus = ceiling(con_perc),
  stringsAsFactors = FALSE
)

# Combine all rows
estuary_table <- rbind(estuary_table, species_row, percent_row)

# View the table
view(estuary_table)
```


```{r}
EF3 <- read_xlsx('EstuaryFishSurveyData.xlsx')
EPA <- read_xlsx('Protected estuaries.xlsx')

# Merge EF and EPA using different column names
matched_df <- merge(EF3, EPA, by.x = "Estuary", by.y = "Protected Estuaries")
```

Replace NA values with 0

```{r}
EF3[is.na(EF3)] <- 0
```
# Quick diagnostic code:
 Check columns in EF
print(colnames(EF))

 If Estuary_clean missing, create it
if (!"Estuary_clean" %in% colnames(EF)) {
  EF$Estuary_clean <- tolower(trimws(EF$Estuary))
}

 Same for EF2
print(colnames(EF2))
if (!"Estuary_clean" %in% colnames(EF2)) {
  EF2$Estuary_clean <- tolower(trimws(EF2$Estuary))
}

```{r}
# Clean estuary names in both datasets
EF3$Estuary_clean <- tolower(trimws(EF3$Estuary))
matched_df$Estuary_clean <- tolower(trimws(matched_df$Estuary))  # assuming matched_df has Estuary column

# Create EPA score as a vector
epa_score <- ifelse(EF3$Estuary_clean %in% matched_df$Estuary_clean, 1, 0)

# Optional: name the vector by estuary for clarity
names(epa_score) <- EF3$Estuary

EF3$EPA_score <- epa_score

EF3 <- EF3 %>%
  left_join(EF[, c("Estuary_clean", "AlphaList", "SW_list", "BZ_list", "Slice80_list", "RC_list", "con_list")],
            by = "Estuary_clean") %>%
  left_join(EF2[, c("Estuary_clean", "comp_list")],
            by = "Estuary_clean")


# Vector of columns you want to fix
cols_to_fix <- c("AlphaList", "SW_list", "BZ_list", "Slice80_list", "RC_list", "comp_list")

# Check which of these columns exist in EF3
cols_present <- cols_to_fix[cols_to_fix %in% colnames(EF3)]

# Replace NA with 0 in existing columns only
EF3 <- EF3 %>% mutate(across(all_of(cols_present), ~replace_na(., 0)))

pro_table <- data.frame(
  Estuary = EF3$Estuary,
  EPA = EF3$EPA_score,
  Alpha = EF3$AlphaList,
  SW = EF3$SW_list,
  BZ = EF3$BZ_list,
  Slice80 = EF3$Slice80_list,
  RC = EF3$RC_list,
  Comp = EF3$comp_list,
    stringsAsFactors = FALSE
)
pro_table[is.na(pro_table)] <- 0


# Make sure the columns are numeric
pro_table$EPA <- as.numeric(pro_table$EPA)
pro_table$Alpha <- as.numeric(pro_table$Alpha)
pro_table$SW <- as.numeric(pro_table$SW)
pro_table$BZ <- as.numeric(pro_table$BZ)
pro_table$Slice80 <- as.numeric(pro_table$Slice80)
pro_table$RC <- as.numeric(pro_table$RC)
pro_table$Comp <- as.numeric(pro_table$Comp)

# Create a new Consensus score by summing across the relevant columns
pro_table$Consensus <- rowSums(pro_table[, c("EPA", "Alpha", "SW", "BZ", "Slice80", "RC", "Comp")])

pro_table <- pro_table[order(-pro_table$Consensus), ]


# View the table
view(pro_table)

```

## Conclusion
Create a table of the lists, the number of species protected and the percentage of species protected


